---
title: "Descriptive Analysis on Geneomic Alterations in 120 JMP_Pac Patients - Obesity at BMI higher than 25"
author: "Yotsawat Pomyen"
output:
  html_notebook: default
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

This is the descriptive analysis on mutational alterations data from Caris patient sequencing reports. The data were extracted from XML by Kasonde using 01.extract_genomic_data.Rmd and processed by using 03.complexheatmap.Rmd. The scripts are in Dang-Lab GitHub Repo under JMP_PAC folder.

```{r helper functions, include=FALSE}
source("~/JMP_Pac_mutation/Dang-Lab/JMP_PAC/02.oncoplot_helpers.R")

# Load libraries
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressWarnings(library(readxl))
suppressWarnings(library(ggplot2))
suppressWarnings(library(forcats))
suppressMessages(library(dplyr))
suppressWarnings(library(tidyverse))
suppressWarnings(library(grid))
suppressWarnings(library(tables))
#suppressWarnings(library(flextable))
suppressWarnings(library(gt))
suppressWarnings(library(janitor))
suppressWarnings(library(ggmosaic))
suppressWarnings(library(cowplot))
suppressWarnings(library(gtExtras))
```

### Data used in the analysis: 

1.  **Summary Report (sreport_sheet3)**: This data set provides a
    detailed report of molecular alterations across genes, summarizing
    the types and frequencies of mutations observed in the study cohort.

2.  **Clinical Dataset (Masterlist)**: Contains clinical data for all
    patients included in the study. This masterlist links genetic data
    with patient identifiers, clinical outcomes, and other relevant
    clinical parameters.

```{r Load data set, include=FALSE}
# Load Mutation Summaries from CARIS REPORTS
setwd("/oceanus/collab/InternalJeff/users/hxd052/data/JMP_PAC/datasets")
sreport_sheet3 <- read_xlsx(path = "/oceanus/collab/InternalJeff/users/hxd052/data/JMP_PAC/datasets/excels/summaries/jmp_pac_mutation_summary_v2.xlsx") # update for final pipeline (mutation summary report)
head(sreport_sheet3)
dim(sreport_sheet3)
colnames(sreport_sheet3)
cat("Mutation Summaries loaded!!!\n")

# Load Main Datasets ----
mst_list <- read_xlsx("/oceanus/collab/InternalJeff/users/hxd052/data/JMP_PAC/datasets/excels/JMP_PaC_NAFLD_n502_Extended_Dataset.xlsx")
head(mst_list)
dim(mst_list)
colnames(mst_list)
cat("Main dataset loaded!!!\n")

# Extract sub-dataframe of patients with sequencing
sequenced <- mst_list %>% dplyr::select(`JMP_PAC ID`, 
                                        `Res_CARIS ID`, 
                                        `Res_SEQ-WTS (Y/N)`, 
                                        `Res_SEQ-WES (T/F)`, 
                                        `NASH/NAFLD Status (Y/N)`,
                                        `Hypertension (Y/N)`,
                                        `Hyperlipidemia Status (Y/N)`,
                                        `BMI (kg/m2) at DOS`,
                                        `DIABETES TYPE (Pre-op)`)
# Filter rows where `Res_CARIS ID` is Y only (patients with WTS sequencing)
wts_sequenced <- sequenced %>% dplyr::filter(`Res_SEQ-WTS (Y/N)` == "Y")
cat("Filtered sequenced patients.\n")
to.drop <- c('Res_SEQ-WTS (Y/N)','Res_SEQ-WES (T/F)')
wts_sequenced <- wts_sequenced %>% select(-any_of(to.drop))
wts_sequenced[which(wts_sequenced$`NASH/NAFLD Status (Y/N)` == 'NA'),3] <- 'N'
```

```{r Change DIABETES TYPE to DM Status, include=FALSE}
wts_sequenced <- wts_sequenced %>% mutate(DM = case_when(`DIABETES TYPE (Pre-op)` == "NA" | `DIABETES TYPE (Pre-op)` == "prediabetic" | `DIABETES TYPE (Pre-op)` == "Prediabetic" | `DIABETES TYPE (Pre-op)` == "unspecified" | `DIABETES TYPE (Pre-op)` == "post-pancreatectomy diabetes" ~ "N",
                               `DIABETES TYPE (Pre-op)` == "type 1" | `DIABETES TYPE (Pre-op)` == "type 2" ~ "Y"))
```

```{r BMI > 25/30 cutoff, include=FALSE}
wts_sequenced$`BMI (kg/m2) at DOS` <- as.numeric(wts_sequenced$`BMI (kg/m2) at DOS`)

wts_sequenced <- wts_sequenced %>% mutate(`Obese30` = case_when(`BMI (kg/m2) at DOS` < 30 ~ 'N', `BMI (kg/m2) at DOS` >= 30 ~ 'Y'))

wts_sequenced <- wts_sequenced %>% mutate(`Obese25` = case_when(`BMI (kg/m2) at DOS` <= 25 ~ 'N', `BMI (kg/m2) at DOS` >= 25 ~ 'Y'))

wts_sequenced
```

```{r MASLD status determination, echo=FALSE}
wts_sequenced <- wts_sequenced %>% mutate(MASLD25 = case_when(`NASH/NAFLD Status (Y/N)` == "Y" & `Hypertension (Y/N)` == "Y" ~ "Y",
                                           `NASH/NAFLD Status (Y/N)` == "Y" & `Hyperlipidemia Status (Y/N)` == "Y" ~ "Y",
                                           `NASH/NAFLD Status (Y/N)` == "Y" & `BMI (kg/m2) at DOS` > 25 ~ "Y",
                                           `NASH/NAFLD Status (Y/N)` == "Y" & DM == "Y" ~ "Y", 
                                           `NASH/NAFLD Status (Y/N)` == "N" ~ "N"))

wts_sequenced <- wts_sequenced %>% mutate(MASLD30 = case_when(`NASH/NAFLD Status (Y/N)` == "Y" & `Hypertension (Y/N)` == "Y" ~ "Y",
                                           `NASH/NAFLD Status (Y/N)` == "Y" & `Hyperlipidemia Status (Y/N)` == "Y" ~ "Y",
                                           `NASH/NAFLD Status (Y/N)` == "Y" & `BMI (kg/m2) at DOS` > 30 ~ "Y",
                                           `NASH/NAFLD Status (Y/N)` == "Y" & DM == "Y" ~ "Y", 
                                           `NASH/NAFLD Status (Y/N)` == "N" ~ "N",
                                           `NASH/NAFLD Status (Y/N)` == "N" | `Hyperlipidemia Status (Y/N)` == "N" | DM == "Y" & `BMI (kg/m2) at DOS` < 30 ~ "N" ))

masld_criteria <- function(x){
  table(wts_sequenced[['NASH/NAFLD Status (Y/N)']],wts_sequenced[[x]])
  }
mc <- c("Hypertension (Y/N)","Hyperlipidemia Status (Y/N)","DM", "Obese25", "Obese30")
## everyone is OBESE so no need to include 
mc_table <- lapply(mc,masld_criteria)
names(mc_table) <- mc
mc.fisher <- lapply(mc_table, fisher.test)
```

### MASLD diagnostic criteria
According to [AASLD](https://www.aasld.org/new-masld-nomenclature), adults with MASLD must have NAFLD (steatosis) and at least one of the four following conditions: 

- Hypertension 
  - BP $\ge$ 130/85 mmHg OR
  - specific antihypertensive drug treatment
  - We used **diagnosis of hypertension** from EMR 

```{r NAFLD v Hypertension, echo=FALSE}
nafld.hypertension <- wts_sequenced %>%
  tabyl(`Obese25`,`Hypertension (Y/N)`) %>%
  gt(
    rowname_col = 'Obese25'
  ) %>%
  tab_spanner(
    columns = 2:3,
    label = "Hypertension"
  ) %>%
  tab_stubhead(
    label = "Obesity25 status"
  ) %>% 
  tab_footnote(
    footnote = paste("Fisher test p-value = ",signif(mc.fisher[["Hypertension (Y/N)"]]$p.value, digits = 3)) 
  )
nafld.hypertension
```

- Hyperlipidemia 
  - Plasma Triglyceride $\ge$ 1.70 mmol [150 mg/dL] OR
  - Plasma HDL-cholesterol $\le$ 1.0 mmol [40 mg/dL] (M)/1.3 mmol [50 mg/dL] (F) OR
  - Lipid lowering treatment
  - We used **diagnosis of hyperlipidemia** from EMR

```{r NAFLD v Hyperlipidemia, echo=FALSE}
wts_sequenced$`Hyperlipidemia Status (Y/N)`[52] <- 'N'
nafld.hyperlipidemia <- wts_sequenced %>%
  tabyl(`Obese25`,`Hyperlipidemia Status (Y/N)`) %>%
  gt(
    rowname_col = 'Obese25'
  ) %>%
  tab_spanner(
    columns = 2:3,
    label = "Hyperlipidemia"
  ) %>%
  tab_stubhead(
    label = "Obesity25 Status"
  ) %>% 
  tab_footnote(
    footnote = paste("Fisher test p-value = ",signif(mc.fisher[["Hyperlipidemia Status (Y/N)"]]$p.value, digits = 3))
  )
nafld.hyperlipidemia
```

- Diabetes
  - Fasting serum glucose $\ge$ 5.6 nmol/L [100 mg/dL]\ OR 
  - 2-hour-post-load glucose levels $\ge$ 7.8 nmol/L OR
  - HbA1c $\ge$ 5.7% [39 nmol/L] OR 
  - Treatment for DM
  - We used **diagnosis of Type 1 or Type 2 DM** from EMR, excluding pre-DM/NA/unspecified 

```{r NAFLD v DM, echo=FALSE}
nafld.dm <- wts_sequenced %>%
  tabyl(`Obese25`,`DM`) %>%
  gt(
    rowname_col = 'Obese25'
  ) %>%
  tab_spanner(
    columns = 2:3,
    label = "Diabetes"
  ) %>%
  tab_stubhead(
    label = "Obesity25 Status"
  ) %>% 
  tab_footnote(
    footnote = paste("Fisher test p-value = ",signif(mc.fisher[["DM"]]$p.value, digits = 3)) 
  )
nafld.dm
```

- Obesity 
  - BMI $\ge$ 25 OR 
  - WC $\ge$ 94 cm(M)/80 cm(F) OR 
  - Ethnicity adjusted 
  - In our analysis we used two criteria: **BMI $\ge$ 25** and **BMI $\ge$ 30**, with this cutoff everyone is obese 

```{r NAFLD v Obese BMI > 25, echo=FALSE}
nafld.obese25 <- wts_sequenced %>%
  tabyl(`NASH/NAFLD Status (Y/N)`,`Obese25`) %>%
  gt(
    rowname_col = 'NASH/NAFLD Status (Y/N)'
  ) %>%
  tab_spanner(
    columns = 2:3,
    label = "Obese (BMI >= 25)"
  ) %>%
  tab_stubhead(
    label = "NASH/NAFLD Status"
  ) %>% 
  tab_footnote(
    footnote = paste("Fisher test p-value = ",signif(mc.fisher[["Obese25"]]$p.value, digits = 3)) 
  )
#nafld.obese25
```

```{r NAFLD v Obese BMI > 30, echo=FALSE}
nafld.obese30 <- wts_sequenced %>%
  tabyl(`NASH/NAFLD Status (Y/N)`,`Obese30`) %>%
  gt(
    rowname_col = 'NASH/NAFLD Status (Y/N)'
  ) %>%
  tab_spanner(
    columns = 2:3,
    label = "Obese (BMI >= 30)"
  ) %>%
  tab_stubhead(
    label = "NASH/NAFLD Status"
  ) %>% 
  tab_footnote(
    footnote = paste("Fisher test p-value = ",signif(mc.fisher[["Obese30"]]$p.value, digits = 3)) 
  )
#nafld.obese30
```
\pagebreak
### MASLD Patient Status

##### MASLD Patient Status with BMI $\ge$ 25
```{r MASLD Patient Status BMI 25, echo=FALSE}
masld25 <- wts_sequenced %>%
  tabyl(`MASLD25`) %>%
  gt(
    rowname_col = 'MASLD25'
  ) %>%
  tab_spanner(
    columns = 1:3,
    label = "MASLD 25"
  ) 
masld25
```

##### MASLD Patient Status with BMI $\ge$ 30
```{r MASLD Patient Status BMI 30, echo=FALSE}
masld30 <- wts_sequenced %>%
  tabyl(`MASLD30`) %>%
  gt(
    rowname_col = 'MASLD30'
  ) %>%
  tab_spanner(
    columns = 1:3,
    label = "MASLD 30"
  )
masld30
```

MASLD patients between BMI $\ge$ 25 or 30 is not different

\pagebreak
### KRAS Mutation v MASLD
```{r Extract KRAS mutation types in long format, echo=FALSE}
kras_mutation <- sreport_sheet3 %>% filter(biomarkerName == 'KRAS' & result_group == 'Mutated')
colnames(kras_mutation)[2] <- 'JMP_PAC ID'
kras_mutation <- as_tibble(merge(kras_mutation, wts_sequenced, by = 'JMP_PAC ID', all = T)) %>% filter(biomarkerName == 'KRAS' & result_group == 'Mutated')

colnames(kras_mutation)[9:10] <- c('KRAS_Mutation','KRAS_Protein')
```

```{r KRAS nucleotide changes plot, echo=FALSE}
kras.coding.gg <- ggplot(data = kras_mutation) + 
  geom_mosaic(aes(x = product(KRAS_Mutation, Obese25), fill = KRAS_Mutation), divider = c('vbar','hspine')) +
  theme_minimal() +
  theme(axis.ticks.x = element_blank(), panel.grid.major = element_blank()) + 
  labs(y = 'KRAS mutation', x = 'Obesity25 Status', title = paste('KRAS nucleotide changes\nFisher test p-value = ',signif(fisher.test(kras_mutation$KRAS_Mutation,kras_mutation$Obese25)$p.value, digits = 3), sep = ''))
kras.coding.gg
```

```{r KRAS protein changes plot, echo=FALSE}
kras.protein <- with(kras_mutation, table(KRAS_Protein,Obese25))

kras.protein.gg <- ggplot(data = kras_mutation) + 
  geom_mosaic(aes(x = product(KRAS_Protein, Obese25), fill = KRAS_Protein), divider = c('vbar','hspine')) +
  theme_minimal() +
  theme(axis.ticks.x = element_blank(), panel.grid.major = element_blank()) + 
  labs(y = 'KRAS AA Changes', x = 'Obesity25 Status', title = paste('KRAS amino acid changes\nFisher test p-value = ',signif(fisher.test(kras.protein)$p.value, digits = 3), sep = ''))
kras.protein.gg
```

```{r}

kras_mutation <- kras_mutation %>% mutate(G12D = case_when(KRAS_Protein == 'p.G12D' ~ 'Y',
          KRAS_Protein != 'p.G12D' ~ 'N'))

kras.G12D <- with(kras_mutation, table(Obese25,G12D))

kras.g12d.gg <- ggplot(data = kras_mutation) + 
  geom_mosaic(aes(x = product(G12D, Obese25), fill = G12D), divider = c('vbar','hspine')) +
  theme_minimal() +
  theme(axis.ticks.x = element_blank(), panel.grid.major = element_blank()) + 
  labs(y = 'KRAS G12D v other changes', x = 'Obesity25 Status', title = paste('KRAS G12D status\nFisher test p-value = ',signif(fisher.test(kras.G12D)$p.value, digits = 3), sep = '')) +
  scale_fill_manual(values = c('darkgreen','darkred'))

kras.g12d.gg
```
\pagebreak
### Other Cancer Type Relevant Biomarkers status

There are three types of alterations reported for each biomarker: 
1. DNA mutation: SNVs, InDels
2. Amplification: CNVs
3. RNA fusion

Therefore, the number of alterations are combination of those types. They will not add up to 120, except for MSI and LoH.

```{r Extract all Cancer Type Relevant Biomarkers, echo=FALSE}
cancer_biomarker <- c('BRAF','Microsatellite Instability (MSI)','NTRK1','RET','Tumor Mutational Burden (TMB)','ATM','BRCA1','BRCA2','KRAS','MTAP','NRG1','PALB2','SMAD4','Genomic Loss of Heterozygosity (LOH)','HLA-A','HLA-B','HLA-C')
```

```{r BRAF table, echo=FALSE}
  mut_table <- sreport_sheet3 %>% filter(biomarkerName == biomarker)
  colnames(mut_table)[2] <- 'JMP_PAC ID'
  mut_table <- as_tibble(merge(mut_table, wts_sequenced, by = 'JMP_PAC ID', all = T) %>% filter(biomarkerName == biomarker))
  # colnames(mut_table)[10] <- cancer_biomarker[1]
  #with(mut_table, table(result_group,MASLD25))
  mut_table %>%
    tabyl(`Obese25`,`result_group`,) %>%
    gt(
      rowname_col = 'Obese25'
    ) %>%
    tab_spanner(
      columns = 2:4,
      label = biomarker
    ) %>%
    tab_stubhead(
      label = 'Obesity25 Status'
    ) %>%
    tab_footnote(
      footnote = paste('Fisher test p-value = ',signif(fisher.test(mut_table$Obese25,mut_table$result_group)$p.value, digits = 3))
      )
```

```{r Microsatellite Instability (MSI) table, echo=FALSE}
  mut_table <- sreport_sheet3 %>% filter(biomarkerName == 'Microsatellite Instability (MSI)')
  colnames(mut_table)[2] <- 'JMP_PAC ID'
  mut_table <- as_tibble(merge(mut_table, wts_sequenced, by = 'JMP_PAC ID', all = T) %>% filter(biomarkerName == 'Microsatellite Instability (MSI)'))
  # colnames(mut_table)[10] <- cancer_biomarker[1]
  #with(mut_table, table(result_group,MASLD25))
  mut_table %>%
    tabyl(`Obese25`,`result_group`,) %>%
    gt(
      rowname_col = 'Obese25'
    ) %>%
    tab_spanner(
      columns = 2:4,
      label = 'Microsatellite Instability'
    ) %>%
    tab_stubhead(
      label = 'Obesity25 Status'
    ) %>%
    tab_footnote(
      footnote = paste('Fisher test p-value = ',signif(fisher.test(mut_table$Obese25,mut_table$result_group)$p.value, digits = 3))
      )
```

```{r NTRK1 table, echo=FALSE}
  mut_table <- sreport_sheet3 %>% filter(biomarkerName == 'NTRK1')
  colnames(mut_table)[2] <- 'JMP_PAC ID'
  mut_table <- as_tibble(merge(mut_table, wts_sequenced, by = 'JMP_PAC ID', all = T) %>% filter(biomarkerName == 'NTRK1'))
  # colnames(mut_table)[10] <- cancer_biomarker[1]
  #with(mut_table, table(result_group,MASLD25))
  mut_table %>%
    tabyl(`Obese25`,`result_group`,) %>%
    gt(
      rowname_col = 'Obese25'
    ) %>%
    tab_spanner(
      columns = 2:4,
      label = 'NTRK1'
    ) %>%
    tab_stubhead(
      label = 'Obesity25 Status'
    ) %>%
    tab_footnote(
      footnote = paste('Fisher test p-value = ',signif(fisher.test(mut_table$Obese25,mut_table$result_group)$p.value, digits = 3))
      )
```
\pagebreak
```{r RET table, echo=FALSE}
  mut_table <- sreport_sheet3 %>% filter(biomarkerName == 'RET')
  colnames(mut_table)[2] <- 'JMP_PAC ID'
  mut_table <- as_tibble(merge(mut_table, wts_sequenced, by = 'JMP_PAC ID', all = T) %>% filter(biomarkerName == 'RET'))
  # colnames(mut_table)[10] <- cancer_biomarker[1]
  #with(mut_table, table(result_group,MASLD25))
  mut_table %>%
    tabyl(`Obese25`,`result_group`,) %>%
    gt(
      rowname_col = 'Obese25'
    ) %>%
    tab_spanner(
      columns = 2:3,
      label = 'RET'
    ) %>%
    tab_stubhead(
      label = 'Obesity25 Status'
    ) %>%
    tab_footnote(
      footnote = paste('Fisher test p-value = ',signif(fisher.test(mut_table$Obese25,mut_table$result_group)$p.value, digits = 3))
      )
```

```{r Tumor Mutational Burden (TMB) table, echo=FALSE}
  mut_table <- sreport_sheet3 %>% filter(biomarkerName == 'Tumor Mutational Burden (TMB)')
  colnames(mut_table)[2] <- 'JMP_PAC ID'
  mut_table <- as_tibble(merge(mut_table, wts_sequenced, by = 'JMP_PAC ID', all = T) %>% filter(biomarkerName == 'Tumor Mutational Burden (TMB)'))
  # colnames(mut_table)[10] <- cancer_biomarker[1]
  #with(mut_table, table(result_group,MASLD25))
  mut_table %>%
    tabyl(`Obese25`,`result_group`,) %>%
    gt(
      rowname_col = 'Obese25'
    ) %>%
    tab_spanner(
      columns = 2:3,
      label = 'Tumor Mutational Burden'
    ) %>%
    tab_stubhead(
      label = 'Obesity25 Status'
    ) %>%
    tab_footnote(
      footnote = paste('Fisher test p-value = ',signif(fisher.test(mut_table$Obese25,mut_table$result_group)$p.value, digits = 3))
      )
```

```{r ATM table, echo=FALSE}
  mut_table <- sreport_sheet3 %>% filter(biomarkerName == 'ATM')
  colnames(mut_table)[2] <- 'JMP_PAC ID'
  mut_table <- as_tibble(merge(mut_table, wts_sequenced, by = 'JMP_PAC ID', all = T) %>% filter(biomarkerName == 'ATM'))
  # colnames(mut_table)[10] <- cancer_biomarker[1]
  #with(mut_table, table(result_group,MASLD25))
  mut_table %>%
    tabyl(`Obese25`,`result_group`,) %>%
    gt(
      rowname_col = 'Obese25'
    ) %>%
    tab_spanner(
      columns = 2:4,
      label = 'ATM'
    ) %>%
    tab_stubhead(
      label = 'Obesity25 Status'
    ) %>%
    tab_footnote(
      footnote = paste('Fisher test p-value = ',signif(fisher.test(mut_table$Obese25,mut_table$result_group)$p.value, digits = 3))
      )
```

```{r BRCA1 table, echo=FALSE}
  mut_table <- sreport_sheet3 %>% filter(biomarkerName == 'BRCA1')
  colnames(mut_table)[2] <- 'JMP_PAC ID'
  mut_table <- as_tibble(merge(mut_table, wts_sequenced, by = 'JMP_PAC ID', all = T) %>% filter(biomarkerName == 'BRCA1'))
  # colnames(mut_table)[10] <- cancer_biomarker[1]
  #with(mut_table, table(result_group,MASLD25))
  mut_table %>%
    tabyl(`Obese25`,`result_group`,) %>%
    gt(
      rowname_col = 'Obese25'
    ) %>%
    tab_spanner(
      columns = 2:3,
      label = 'BRCA1'
    ) %>%
    tab_stubhead(
      label = 'Obesity25 Status'
    ) %>%
    tab_footnote(
      footnote = paste('Fisher test p-value = ',signif(fisher.test(mut_table$Obese25,mut_table$result_group)$p.value, digits = 3))
      )
```
\pagebreak
```{r BRCA2 table, echo=FALSE}
  mut_table <- sreport_sheet3 %>% filter(biomarkerName == 'BRCA2')
  colnames(mut_table)[2] <- 'JMP_PAC ID'
  mut_table <- as_tibble(merge(mut_table, wts_sequenced, by = 'JMP_PAC ID', all = T) %>% filter(biomarkerName == 'BRCA2'))
  # colnames(mut_table)[10] <- cancer_biomarker[1]
  #with(mut_table, table(result_group,MASLD25))
  mut_table %>%
    tabyl(`Obese25`,`result_group`,) %>%
    gt(
      rowname_col = 'Obese25'
    ) %>%
    tab_spanner(
      columns = 2:4,
      label = 'BRCA2'
    ) %>%
    tab_stubhead(
      label = 'Obesity25 Status'
    ) %>%
    tab_footnote(
      footnote = paste('Fisher test p-value = ',signif(fisher.test(mut_table$Obese25,mut_table$result_group)$p.value, digits = 3))
      )
```

```{r MTAP table, echo=FALSE}
  mut_table <- sreport_sheet3 %>% filter(biomarkerName == 'MTAP')
  colnames(mut_table)[2] <- 'JMP_PAC ID'
  mut_table <- as_tibble(merge(mut_table, wts_sequenced, by = 'JMP_PAC ID', all = T) %>% filter(biomarkerName == 'MTAP'))
  # colnames(mut_table)[10] <- cancer_biomarker[1]
  #with(mut_table, table(result_group,MASLD25))
  mut_table %>%
    tabyl(`Obese25`,`result_group`,) %>%
    gt(
      rowname_col = 'Obese25'
    ) %>%
    tab_spanner(
      columns = 2:4,
      label = 'MTAP'
    ) %>%
    tab_stubhead(
      label = 'Obesity25 Status'
    ) %>%
    tab_footnote(
      footnote = paste('Fisher test p-value = ',signif(fisher.test(mut_table$Obese25,mut_table$result_group)$p.value, digits = 3))
      )
```

```{r NRG1 table, echo=FALSE}
  mut_table <- sreport_sheet3 %>% filter(biomarkerName == 'NRG1')
  colnames(mut_table)[2] <- 'JMP_PAC ID'
  mut_table <- as_tibble(merge(mut_table, wts_sequenced, by = 'JMP_PAC ID', all = T) %>% filter(biomarkerName == 'NRG1'))
  # colnames(mut_table)[10] <- cancer_biomarker[1]
  #with(mut_table, table(result_group,MASLD25))
  mut_table %>%
    tabyl(`Obese25`,`result_group`,) %>%
    gt(
      rowname_col = 'Obese25'
    ) %>%
    tab_spanner(
      columns = 2:3,
      label = 'NRG1'
    ) %>%
    tab_stubhead(
      label = 'Obesity25 Status'
    ) %>%
    tab_footnote(
      footnote = paste('Fisher test p-value = ',signif(fisher.test(mut_table$Obese25,mut_table$result_group)$p.value, digits = 3))
      )
```

```{r PALB2 table, echo=FALSE}
  mut_table <- sreport_sheet3 %>% filter(biomarkerName == 'PALB2')
  colnames(mut_table)[2] <- 'JMP_PAC ID'
  mut_table <- as_tibble(merge(mut_table, wts_sequenced, by = 'JMP_PAC ID', all = T) %>% filter(biomarkerName == 'PALB2'))
  # colnames(mut_table)[10] <- cancer_biomarker[1]
  #with(mut_table, table(result_group,MASLD25))
  mut_table %>%
    tabyl(`Obese25`,`result_group`,) %>%
    gt(
      rowname_col = 'Obese25'
    ) %>%
    tab_spanner(
      columns = 2:3,
      label = 'PALB2'
    ) %>%
    tab_stubhead(
      label = 'Obesity25 Status'
    ) %>%
    tab_footnote(
      footnote = paste('Fisher test p-value = ',signif(fisher.test(mut_table$Obese25,mut_table$result_group)$p.value, digits = 3))
      )
```
\pagebreak
```{r SMAD4 table, echo=FALSE}
  mut_table <- sreport_sheet3 %>% filter(biomarkerName == 'SMAD4')
  colnames(mut_table)[2] <- 'JMP_PAC ID'
  mut_table <- as_tibble(merge(mut_table, wts_sequenced, by = 'JMP_PAC ID', all = T) %>% filter(biomarkerName == 'SMAD4'))
  # colnames(mut_table)[10] <- cancer_biomarker[1]
  #with(mut_table, table(result_group,MASLD25))
smad4.table <- mut_table %>%
    tabyl(`Obese25`,`result_group`,) %>%
    gt(
      rowname_col = 'Obese25'
    ) %>%
    tab_spanner(
      columns = 2:4,
      label = 'SMAD4'
    ) %>%
    tab_stubhead(
      label = 'Obesity25 Status'
    ) %>%
    tab_footnote(
      footnote = paste('Fisher test p-value = ',signif(fisher.test(mut_table$Obese25,mut_table$result_group)$p.value, digits = 3))
      )
```

```{r Genomic Loss of Heterozygosity (LOH) table, echo=FALSE}
  mut_table <- sreport_sheet3 %>% filter(biomarkerName == 'Genomic Loss of Heterozygosity (LOH)')
  colnames(mut_table)[2] <- 'JMP_PAC ID'
  mut_table <- as_tibble(merge(mut_table, wts_sequenced, by = 'JMP_PAC ID', all = T) %>% filter(biomarkerName == 'Genomic Loss of Heterozygosity (LOH)'))
  # colnames(mut_table)[10] <- cancer_biomarker[1]
  #with(mut_table, table(result_group,MASLD25))
loh.table <- mut_table %>%
    tabyl(`Obese25`,`result_group`,) %>%
    gt(
      rowname_col = 'Obese25'
    ) %>%
    tab_spanner(
      columns = 2:4,
      label = 'Genomic LoH'
    ) %>%
    tab_stubhead(
      label = 'Obesity25 Status'
    ) %>%
    tab_footnote(
      footnote = paste('Fisher test p-value = ',signif(fisher.test(mut_table$Obese25,mut_table$result_group)$p.value, digits = 3))
      )
```

```{r SMAD4 and LoH ,echo=FALSE}
test <- list(smad4.table,loh.table)
gt_two_column_layout(test)
```


```{r HLA-A table, echo=FALSE}
  mut_table <- sreport_sheet3 %>% filter(biomarkerName == 'HLA-A')
  colnames(mut_table)[2] <- 'JMP_PAC ID'
  mut_table <- as_tibble(merge(mut_table, wts_sequenced, by = 'JMP_PAC ID', all = T) %>% filter(biomarkerName == 'HLA-A'))
  # colnames(mut_table)[10] <- cancer_biomarker[1]
  #with(mut_table, table(result_group,MASLD25))
  mut_table %>%
    tabyl(`Obese25`,`result_group`,) %>%
    gt(
      rowname_col = 'Obese25'
    ) %>%
    tab_spanner(
      columns = 2:3,
      label = 'HLA-A'
    ) %>%
    tab_stubhead(
      label = 'Obesity25 Status'
    ) %>%
    tab_footnote(
      footnote = paste('Fisher test p-value = ',signif(fisher.test(mut_table$Obese25,mut_table$result_group)$p.value, digits = 3))
      )
```

#### There is no mutation in HLA-B and HLA-C. 


```{r 2Extract all Cancer Type Relevant Biomarkers, echo=FALSE}
cancer_biomarker <- c('BRAF','Microsatellite Instability (MSI)','NTRK1','RET','Tumor Mutational Burden (TMB)','ATM','BRCA1','BRCA2','KRAS','MTAP','NRG1','PALB2','SMAD4','Genomic Loss of Heterozygosity (LOH)','HLA-A','HLA-B','HLA-C')
```

```{r SMAD4 mutation, eval=FALSE, include=FALSE}
smad4_mutation <- sreport_sheet3 %>% filter(biomarkerName == 'SMAD4' & result_group == 'Mutated')
colnames(smad4_mutation)[2] <- 'JMP_PAC ID'
smad4_mutation <- as_tibble(merge(smad4_mutation, wts_sequenced, by = 'JMP_PAC ID', all = T)) %>% filter(biomarkerName == 'SMAD4' & result_group == 'Mutated')

colnames(smad4_mutation)[9:10] <- c('SMAD4_coding','SMAD4_Protein')
#smad4.coding <- with(smad4_mutation, table(SMAD4_coding,MASLD25))

#smad4.protein <- with(smad4_mutation, table(SMAD4_Protein,MASLD25))

hlaa_mutation <- sreport_sheet3 %>% filter(biomarkerName == 'HLA-A' & result_group == 'Mutated')
colnames(hlaa_mutation)[2] <- 'JMP_PAC ID'
hlaa_mutation <- as_tibble(merge(hlaa_mutation, wts_sequenced, by = 'JMP_PAC ID', all = T)) %>% filter(biomarkerName == 'HLA-A' & result_group == 'Mutated')

#hlaa.class <- with(hlaa_mutation, table(mhClass,MASLD25))
#hlaa.genotype <- with(hlaa_mutation, table(genotype,MASLD25))
```


```{r code not working, eval=FALSE, include=FALSE}
ctrb <- list()
ctrb_table <- function(biomarker){
  mut_table <- sreport_sheet3 %>% filter(biomarkerName == biomarker)
  colnames(mut_table)[2] <- 'JMP_PAC ID'
  mut_table <- as_tibble(merge(mut_table, wts_sequenced, by = 'JMP_PAC ID', all = T) %>% filter(biomarkerName == biomarker))
  # colnames(mut_table)[10] <- cancer_biomarker[1]
  with(mut_table, table(result_group,MASLD25))
}
ctrb <- lapply(cancer_biomarker,ctrb_table)
names(ctrb) <- cancer_biomarker





for (i in 1:length(ctrb)){
  wts_sequenced %>%
  tabyl(`MASLD25`,names(ctrb[i])) %>%
  gt(
    rowname_col = 'MASLD25'
  ) %>%
  tab_spanner(
    columns = 1:2,
    label = x
  ) %>%
  tab_stubhead(
    label = "MASLD Status"
  ) %>% 
  tab_footnote(
    footnote = paste("Fisher test p-value = ",signif(fisher.test(ctrb[[i]])$p.value, digits = 3)) 
  )
}
```
