---
title: "Analysis of RNA-seq Data EV/KO"
author: "Kasonde Chewe"
file: "01.RNA_sgEV_sgNELFE_boxplots_g1andg2_up_down.Rmd"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_notebook
---

# Description
Notebook includes the analysis workflow for RNA-seq, HEP3B sgEV and sgNELFE knockout experiment.
The analysis includes, fold change (log2FC) computation, statisical analysis and boxplot visualization. 
A higher quality render is provided through graphpad prism!
An accompanying, single executable R script can be found for publication 
at XXXX. Analysis references and environment configuration is included at the 
end of the following notebook.


```{r CLEAR ENV VARIABLES, echo=FALSE, include=FALSE}
# clear all environment variables
rm(list=ls())
gc() # call garbage collector 
options(digits = 5)
```





```{r LOAD LIBRARIES AND HELPERS, include=TRUE, echo=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(writexl)
library(ggpubr)
```


```{r LOAD CSV FILES AND GROUPS,  include=TRUE, echo=FALSE, message=FALSE}
dataset_dir <- "~/Projects/Condensate_Paper/02_RNASEQ_FINAL/datasets"
peaks_dir <- "~/Projects/Condensate_Paper/01_GROSEQ_FINAL/peaks"

# read in datasets
rna_counts <- read.csv(paste0(dataset_dir,"/countsMatrix_RNA-sgEV-sgKO_with_genes.csv"))
groupI <- read.delim(paste0(peaks_dir,"/groupI_genes.bed"), header = FALSE)
groupII <- read.delim(paste0(peaks_dir,"/groupII_genes.bed"),  header = FALSE)

```




```{r COUNT MATRIX PREPROCESSING & EDA, include=TRUE, echo=FALSE, message=FALSE}

library(ggprism)
# Summarize counts for duplicate gene names by taking the sum
rna.anno.sum.counts.matrix <- rna_counts %>%
  group_by(Gene.Name) %>%
  summarise(across(everything(), sum, .names = "sum_{.col}")) %>%
  rename_with(~ sub("^sum_", "", .))
dim(rna.anno.sum.counts.matrix)
head(rna.anno.sum.counts.matrix)

# Compute the average expression for each triplicate
# length is 7921 genes
rna.anno.sum.merged.counts.matrix <- rna.anno.sum.counts.matrix %>%
  mutate(
    EV_mean = rowMeans(dplyr::select(., "EV.R1", "EV.R2", "EV.R3")),
    KO_mean = rowMeans(dplyr::select(., "KO.R1", "KO.R2", "KO.R3"))
  )


# Prefiltering exclude low expression genes 
# length is 7225 genes
rna.prefiltered.counts.matrix <- rna.anno.sum.merged.counts.matrix %>%
  filter(Gene.Name != "") %>%
  filter(EV_mean >= 1 | KO_mean >= 1) %>% drop_na()

# Subset matrix for all gene name and averaged, filtered columns 
rna.counts.ev.ko <- as.data.frame(rna.anno.sum.merged.counts.matrix[,c("Gene.Name", "EV_mean", "KO_mean")])
dim(rna.counts.ev.ko)
head(rna.counts.ev.ko)

# # save dataframe copy 
# write_xlsx(rna.counts.ev.ko, "~/Projects/Condensate_Paper/02_RNASEQ_FINAL/datasets/merged_countsMatrix_RNA-sgEV-sgKO_with_genes.xlsx")

# Compute log2FC(sgNELFE/sgEV) 
# Set rownames to Gene.Name and remove Gene.Name column
rna.counts.ev.ko$log2FC <- log2(rna.counts.ev.ko$KO_mean + 1) - log2(rna.counts.ev.ko$EV_mean + 1)
```


```{r}
# GroupI 
# Subset genes in rna.counts.ev.ko by common gene in groupI$V5
groupI_genes <- groupI$V5
groupI_counts <- rna.counts.ev.ko %>%
  filter(Gene.Name %in% groupI_genes) %>% drop_na()

groupI_counts$sgEV <- log2(groupI_counts$EV_mean + 0.01)
groupI_counts$sgNELFE <- log2(groupI_counts$KO_mean + 0.01)


groupI_counts_up <- subset(groupI_counts, log2FC > 0)
groupI_counts_down <- subset(groupI_counts, log2FC < 0)

# GroupII
# Subset genes in rna.counts.ev.ko by common gene in groupII$V4
groupII_genes <- groupII$V4
groupII_counts <- rna.counts.ev.ko %>%
  filter(Gene.Name %in% groupII_genes) %>% drop_na()


groupII_counts$sgEV <- log2(groupII_counts$EV_mean + 0.01)
groupII_counts$sgNELFE <- log2(groupII_counts$KO_mean + 0.01) 

groupII_counts_up <- subset(groupII_counts, log2FC > 0)
groupII_counts_down <- subset(groupII_counts, log2FC < 0)

# Convert data to long format for Group I
groupI_counts_long <- groupI_counts %>%
  pivot_longer(cols = c("sgEV", "sgNELFE"),
               names_to = "Condition",
               values_to = "Expression")

# Convert data to long format for Group I
groupI_counts_up_long <- groupI_counts_up %>%
  pivot_longer(cols = c("sgEV", "sgNELFE"),
               names_to = "Condition",
               values_to = "Expression")

# Convert data to long format for Group I
groupI_counts_down_long <- groupI_counts_down %>%
  pivot_longer(cols = c("sgEV", "sgNELFE"),
               names_to = "Condition",
               values_to = "Expression")


# Convert data to long format for Group II
groupII_counts_long <- groupII_counts %>%
  pivot_longer(cols = c("sgEV", "sgNELFE"),
               names_to = "Condition",
               values_to = "Expression")


# Convert data to long format for Group I
groupII_counts_up_long <- groupII_counts_up %>%
  pivot_longer(cols = c("sgEV", "sgNELFE"),
               names_to = "Condition",
               values_to = "Expression")

# Convert data to long format for Group I
groupII_counts_down_long <- groupII_counts_down %>%
  pivot_longer(cols = c("sgEV", "sgNELFE"),
               names_to = "Condition",
               values_to = "Expression")

```



```{r}
# statistical tests 
# groupI 
g1u_pvalue <- compare_means(Expression ~ Condition, data=groupI_counts_up_long, method = "t.test")
print("\n  Group I Upregulated Genes -- \n T-test statistics \n")
print(as.data.frame(g1u_pvalue))

g1d_pvalue <- compare_means(Expression ~ Condition, data=groupI_counts_down_long, method = "t.test")
g1d_pvalue

# groupI 
g2u_pvalue <- compare_means(Expression ~ Condition, data=groupII_counts_up_long, method = "t.test", p.adjust.method = "BH")
g2d_pvalue <- compare_means(Expression ~ Condition, data=groupII_counts_down_long, method = "t.test")
g2u_pvalue
g2d_pvalue


```


```{r, fig.height=9, fig.width=8}
options( digits = 6)
# Group I Violin PLots 
# up
g1u = length(groupI_counts_up_long$Gene.Name)

pdf("~/Projects/Condensate_Paper/02_RNASEQ_FINAL/plots/violinplots/rnaseq_violinPlots_groupI_upregulated_sgEV_sgNELFE.pdf")
ggplot(groupI_counts_up_long, aes(x = Condition, y = Expression , fill = Condition)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.2)+
  labs(title = paste0("RNAseq Group I Upregulated Genes, n=", g1u ),
       x = "Condition",
       y = "log2(CPM)") +
  scale_fill_manual(values = c("sgEV" = "blue", "sgNELFE" = "red")) + 
  theme_prism(base_size = 18) + 
  stat_compare_means(method = "t.test", paired = FALSE, label = "p.format") 
dev.off()

g1d = length(groupI_counts_down_long$Gene.Name)
pdf("~/Projects/Condensate_Paper/02_RNASEQ_FINAL/plots/violinplots/rnaseq_violinplots_groupI_downregulated_sgEV_sgNELFE.pdf")
ggplot(groupI_counts_down_long, aes(x = Condition, y = Expression, fill = Condition)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.2) +
  labs(title = paste0("RNAseq Group I Downregulated Genes, n=", g1d ),
       x = "Condition",
       y = "log2(CPM)") +
  scale_fill_manual(values = c("sgEV" = "blue", "sgNELFE" = "red")) + 
  theme_prism(base_size = 18) + 
  stat_compare_means(method = "t.test", paired = FALSE, label = "p.format")  # or paired = TRUE if it is paired data
dev.off()
```


```{r, fig.height=9, fig.width=8}
# Group II Violin PLots 
# up
g2u = length(groupII_counts_up_long$Gene.Name)
pdf("~/Projects/Condensate_Paper/02_RNASEQ_FINAL/plots/violinplots/rnaseq_violinPlots_groupII_upregulated_sgEV_sgNELFE.pdf")
ggplot(groupII_counts_up_long, aes(x = Condition, y = Expression , fill  = Condition)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.2)+
  theme_minimal() +
  labs(title = paste0("RNAseq Group II Upregulated Genes, n=", g2u ),
       x = "Condition",
       y = "log2(CPM)") +
  scale_fill_manual(values = c("sgEV" = "blue", "sgNELFE" = "red")) + 
  theme_prism(base_size = 18) + 
  stat_compare_means(method = "t.test", paired = FALSE, label = "p.format") 
dev.off()

# down 
g2d = length(groupII_counts_down_long$Gene.Name)
pdf("~/Projects/Condensate_Paper/02_RNASEQ_FINAL/plots/violinplots/rnaseq_violinplots_groupII_downregulated_sgEV_sgNELFE.pdf")
ggplot(groupII_counts_down_long, aes(x = Condition, y = Expression, fill = Condition)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.2)+
  theme_minimal() +
  labs(title = paste0("RNAseq Group II Downregulated Genes, n=", g2d ),
       x = "Condition",
       y = "log2(CPM)") +
  scale_fill_manual(values = c("sgEV" = "blue", "sgNELFE" = "red")) + 
  theme_prism(base_size = 18) + 
  stat_compare_means(method = "t.test", paired = FALSE, label = "p.format")
dev.off()
```



```{r}
# Create the boxplot for Group I
ggplot(groupI_counts_up_long, aes(x = Condition, y = log2(Expression+0.1) , color = Condition)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Boxplot of EV and KO Mean Expressions for Group I Genes",
       x = "Condition",
       y = "log2(CPM)") +
  scale_fill_manual(values = c("EV_mean" = "blue", "KO_mean" = "red")) + 
  theme_prism(base_size = 18)

# Create the boxplot for Group II
ggplot(groupII_counts_long, aes(x = Condition, y = log2(Expression+0.1), color = Condition)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Boxplot of EV and KO Mean Expressions for Group II Genes",
       x = "Condition",
       y = "log2(CPM)") +
  scale_fill_manual(values = c("EV_mean" = "blue", "KO_mean" = "red")) + 
  theme_prism(base_size = 18)
```




```{r}
writexl::write_xlsx(groupI_counts, "~/Projects/Condensate_Paper/02_RNASEQ_FINAL/datasets/group1.xlsx")

writexl::write_xlsx(groupI_counts_up, "~/Projects/Condensate_Paper/02_RNASEQ_FINAL/datasets/group1_up.xlsx")
writexl::write_xlsx(groupI_counts_down, "~/Projects/Condensate_Paper/02_RNASEQ_FINAL/datasets/group1_down.xlsx")


writexl::write_xlsx(groupII_counts, "~/Projects/Condensate_Paper/02_RNASEQ_FINAL/datasets/group2.xlsx")
writexl::write_xlsx(groupII_counts_up, "~/Projects/Condensate_Paper/02_RNASEQ_FINAL/datasets/group2_up.xlsx")
writexl::write_xlsx(groupII_counts_down, "~/Projects/Condensate_Paper/02_RNASEQ_FINAL/datasets/group2_down.xlsx")

```

